Bayesian neural networks for sparse coding

### 2018-10-27 icassp_2019/paper.tex
minor changes after proof-read in icassp paper

\documentclass{article}
\usepackage{spconf,amsmath,graphicx}
\usepackage{mathtools}
\usepackage{xfrac}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{times}
\usepackage{float}
\usepackage{algorithm, algorithmic}

@@

@@

\section{Neural networks for sparse coding}
  \label{sec:bayesian_lista}
  This section first reviews existing NNs for sparse coding and then describes the novel Bayesian NN.

\subsection{Frequentist neural networks}  
\label{subsec:nn_sc}
  The NN approach to sparse coding is based on earlier Iterative Shrinkage and Thresholding Algorithm (\textsc{ista}) \cite{daubechies2004iterative}. It addresses the sparse coding problem as the linear regression problem with the $l1$ penalty that promotes sparsity. For the linear regression model with observations $\mathbf{y} \in \mathbb{R}^K$, the design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$, and the sparse unknown vector of weights $\boldsymbol\beta \in \mathbb{R}^D$, \textsc{ista} minimises
  \begin{equation}
  \label{eq:regression_problem}
  ||\mathbf{X}\boldsymbol\beta - \mathbf{y}||_2^2 + \alpha ||\boldsymbol\beta||_1
  \end{equation}
w.r.t.  $\boldsymbol\beta$,  where $\alpha$ is the penalty parameter.

@@

  In contrast to \textsc{ista}, Learned \textsc{ista} (\textsc{lista}) \cite{gregor2010learning} learns the values of matrices $\mathbf{W}$ and $\mathbf{S}$ based on a set of pairs $\{\mathbf{Y}, \mathbf{B}\}=\{\mathbf{y}^{(n)}, \boldsymbol\beta^{(n)}\}_{n=1}^N$, where $N$ is the number of these pairs. To achieve this, the \textsc{ista} is limited with the fixed amount of iterations $L$ and interpreted as a recurrent NN: every iteration $l$ of \textsc{ista} corresponds to the layer $l$ of \textsc{lista}. A vector $\boldsymbol\beta$ for an observation~$\mathbf{y}$ is predicted by Algorithm~\ref{alg:lista}.

@@

  \subsection{Bayes\textsc{lista}}
  \label{subsec:bayesian_lista}
  This section introduces the proposed Bayesian version of \textsc{lista}. The prior distributions are imposed on the unknown weights
  \begin{equation}
  \label{eq:ws}
  \begin{split}
  p(\mathbf{W}) &= \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} ; 0, \eta^{-1}), \\
  p(\mathbf{S}) &= \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} ; 0, \eta^{-1}),
  \end{split}
  \end{equation}
  where $\eta$ is the precision of the Gaussian distribution.

  For every layer $l$ of Bayes\textsc{lista}, $\widehat{\boldsymbol\beta}_{l}$ is assumed to have the spike and slab distribution with the spike probability $\boldsymbol\omega$, the slab mean $\mathbf{m}$, and the slab variance $\mathbf{v}$
  \begin{equation}
  [\widehat{\boldsymbol\beta}_{l}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
  \end{equation}
  where $\delta_0$ is the delta-function that represents a spike, $[\cdot]_d$ denotes the $d$-th component of a vector. Section \ref{sec:fprop} shows that the output of the next layer $\widehat{\boldsymbol\beta}_{l+1}$ can be approximated with a spike and slab distribution and, therefore, the output of the Bayes\textsc{lista} network $\widehat{\boldsymbol\beta}$ has the spike and slab distribution.

@@

  \section{Uncertainty propagation through soft-thresholding}
  \label{sec:fprop}
  This section describes modification of \textsc{lista} forward propagation (Algorithm~\ref{alg:lista}) to include probability distributions of the random variables introduced in section~\ref{sec:bayesian_lista}.

@@

  Prior factors (\ref{eq:ws}) and (\ref{eq:gamma_eta}) from $p$ are incorporated into $q$ with the EP algorithm~\cite{hernandez2015probabilistic}, i.e. they replace the corresponding approximating factors from $q$, and then $q$ is updated to minimise the Kullback–Leibler divergence.

@@

   \begin{figure*}[!t]
  \centering
  \subfloat[Synthetic for different $L$]{\includegraphics[width=0.24\textwidth]{graphics/synthetic_number_of_layers/nmse_validation}
  \label{fig:nmse_n_layers_synthetic}}
  \hfil
  \subfloat[Synthetic for different $K$]{\includegraphics[width=0.24\textwidth]{graphics/synthetic_undersampling/nmse_validation}
  \label{fig:nmse_undersampling_synthetic}}
  \hfil
  \subfloat[\textsc{mnist} for $K = 100$]{\includegraphics[width=0.24\textwidth]{graphics/mnist/100_nmse_valid}
  \label{fig:nmse_k_100}}
  \hfil
  \subfloat[\textsc{mnist} for $K = 250$]{\includegraphics[width=0.24\textwidth]{graphics/mnist/250_nmse_valid}
  \label{fig:nmse_k_250}}
  \caption{\textsc{nmse} results. The synthetic data for different number of layers~\protect\subref{fig:nmse_n_layers_synthetic} and for different sizes of observations~\protect\subref{fig:nmse_undersampling_synthetic}. \textsc{mnist} results for increasing number of iterations with the observation size $K = 100$~\protect\subref{fig:nmse_k_100} and $K = 250$~\protect\subref{fig:nmse_k_250}}
  \label{fig:number_of_layers_synthetic}
  \end{figure*}

    \begin{figure}[!t]
  \centering
  \includegraphics[width=0.24\textwidth]{graphics/active_mnist/nmse_validation}
  %\subfloat[F measure]{\includegraphics[width=0.48\columnwidth]{graphics/active_mnist/f_measure_validation}}
  \caption{\textsc{nmse} for active learning experiment on \textsc{mnist} data. }
  \label{fig:active_learning_mnist}
  \end{figure}

@@

  \subsection{Predictive performance on \textsc{mnist} data}
  In this experiment, the methods are evaluated on \textsc{mnist} dataset \cite{lecun1998gradient}. The dataset contains images of handwritten digits of size $28 \times 28 = 784$. The design matrix $\mathbf{X}$ is standard random Gaussian. Observations are generated as $\mathbf{y} = \mathbf{X}\boldsymbol\beta$, where $\boldsymbol\beta \in \mathbb{R}^{784}$ are flattened images. We use $100$ images for training and $100$ for test. The shrinkage parameter $\lambda$ is~$0.1$.

@@

  \textsc{mnist} dataset with the observation size $K=100$ is employed. We use the training data is of size $50$, the pool data of size $500$,  and the test data of size $100$. The algorithm learns on the training data and it is evaluated on the test data. To actively collect a next data point from the pool, the algorithm is used to predict a point with the highest uncertainty. The selected point is moved from the pool to the training data and the algorithms learns on the updated training data. Overall, $10$ pool additions are performed. After every addition the performance is measured on the test data. We compare the active approach of adding new points from the pool with the random approach that picks a new data point from the pool at random. The procedure is repeated for $20$ times.

>>>>

\documentclass{article}
\usepackage{spconf,amsmath,graphicx}
\usepackage{mathtools}
\usepackage{xfrac}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{times}
\usepackage{float}
\usepackage{algorithm,algorithmic}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

\section{Neural networks for sparse coding}
  \label{sec:bayesian_lista}
  This section first reviews existing NNs for sparse coding and then introduces the novel Bayesian NN.

\subsection{Frequentist neural networks}  
\label{subsec:nn_sc}
  The NN approach to sparse coding is based on earlier Iterative Shrinkage and Thresholding Algorithm (\textsc{ista}) \cite{daubechies2004iterative}. It addresses the sparse coding problem as the linear regression problem with the $l1$ penalty that promotes sparsity. For the linear regression model with observations $\mathbf{y} \in \mathbb{R}^K$, the design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$, and the sparse unknown vector of weights $\boldsymbol\beta \in \mathbb{R}^D$, \textsc{ista} minimises
  \begin{equation}
  \label{eq:regression_problem}
  ||\mathbf{X}\boldsymbol\beta - \mathbf{y}||_2^2 + \alpha ||\boldsymbol\beta||_1
  \end{equation}
w.r.t.  $\boldsymbol\beta$,  where $\alpha$ is a penalty parameter.

@@

  In contrast to \textsc{ista}, Learned \textsc{ista} (\textsc{lista}) \cite{gregor2010learning} learns the values of matrices $\mathbf{W}$ and $\mathbf{S}$ based on a set of pairs $\{\mathbf{Y}, \mathbf{B}\}=\{\mathbf{y}^{(n)}, \boldsymbol\beta^{(n)}\}_{n=1}^N$, where $N$ is the number of these pairs. To achieve this, \textsc{ista} is limited with the fixed amount of iterations $L$ and interpreted as a recurrent NN: every iteration $l$ of \textsc{ista} corresponds to the layer $l$ of \textsc{lista}. A vector $\boldsymbol\beta$ for an observation~$\mathbf{y}$ is predicted by Algorithm~\ref{alg:lista}.

@@

  \subsection{Bayes\textsc{lista}}
  \label{subsec:bayesian_lista}
  This section introduces the proposed Bayesian version of \textsc{lista} (Bayes\textsc{lista}). The prior distributions are imposed on the unknown weights
  \begin{equation}
  \label{eq:ws}
  \begin{split}
  p(\mathbf{W}) &= \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} ; 0, \eta^{-1}), \\
  p(\mathbf{S}) &= \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} ; 0, \eta^{-1}),
  \end{split}
  \end{equation}
  where $\eta$ is the precision of the Gaussian distribution.

  For every layer $l$ of Bayes\textsc{lista}, $\widehat{\boldsymbol\beta}_{l}$ is assumed to have the spike and slab distribution with the spike probability $\boldsymbol\omega$, the slab mean $\mathbf{m}$, and the slab variance $\mathbf{v}$
  \begin{equation}
  [\widehat{\boldsymbol\beta}_{l}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
  \end{equation}
  where $\delta_0$ is the delta-function that represents a spike, $[\cdot]_d$ denotes the $d$-th component of a vector. Section \ref{sec:fprop} shows that the output of the next layer $\widehat{\boldsymbol\beta}_{l+1}$ can be approximated with the spike and slab distribution and, therefore, the output of the Bayes\textsc{lista} network $\widehat{\boldsymbol\beta}$ has the spike and slab distribution.

@@

  \section{Uncertainty propagation through soft-thresholding}
  \label{sec:fprop}
  This section describes modification of \textsc{lista} forward propagation (Algorithm~\ref{alg:lista}) to include probability distributions of the random variables introduced in section~\ref{subsec:bayesian_lista}.

@@

  Prior factors (\ref{eq:ws}) and (\ref{eq:gamma_eta}) from $p$ are incorporated into $q$ with the EP algorithm~\cite{hernandez2015probabilistic}, i.e. they replace the corresponding approximating factors from $q$, and then $q$ is updated to minimise the Kullback--Leibler divergence.

@@

   \begin{figure*}[!t]
  \centering
  \subfloat[Synthetic for different $L$]{\includegraphics[width=0.24\textwidth]{graphics/synthetic_number_of_layers/nmse_validation}
  \label{fig:nmse_n_layers_synthetic}}
  \hfil
  \subfloat[Synthetic for different $K$]{\includegraphics[width=0.24\textwidth]{graphics/synthetic_undersampling/nmse_validation}
  \label{fig:nmse_undersampling_synthetic}}
  \hfil
  \subfloat[\textsc{mnist} for $K = 100$]{\includegraphics[width=0.24\textwidth]{graphics/mnist/100_nmse_valid}
  \label{fig:nmse_k_100}}
  \hfil
  \subfloat[\textsc{mnist} for $K = 250$]{\includegraphics[width=0.24\textwidth]{graphics/mnist/250_nmse_valid}
  \label{fig:nmse_k_250}}
  \caption{\textsc{nmse} results. The synthetic data results for different number of layers~\protect\subref{fig:nmse_n_layers_synthetic} and for different sizes of observations~\protect\subref{fig:nmse_undersampling_synthetic}. The \textsc{mnist} results for increasing number of iterations with the observation size $K = 100$~\protect\subref{fig:nmse_k_100} and $K = 250$~\protect\subref{fig:nmse_k_250}}
  \label{fig:number_of_layers_synthetic}
  \end{figure*}

    \begin{figure}[!t]
  \centering
  \includegraphics[width=0.24\textwidth]{graphics/active_mnist/nmse_validation}
  %\subfloat[F measure]{\includegraphics[width=0.48\columnwidth]{graphics/active_mnist/f_measure_validation}}
  \caption{\textsc{nmse} for the active learning experiment on \textsc{mnist} data. }
  \label{fig:active_learning_mnist}
  \end{figure}

@@

  \subsection{Predictive performance on \textsc{mnist} data}
  In this experiment, the methods are evaluated on the \textsc{mnist} dataset \cite{lecun1998gradient}. The dataset contains images of handwritten digits of size $28 \times 28 = 784$. The design matrix $\mathbf{X}$ is standard random Gaussian. Observations are generated as $\mathbf{y} = \mathbf{X}\boldsymbol\beta$, where $\boldsymbol\beta \in \mathbb{R}^{784}$ are flattened images. We use $100$ images for training and $100$ for test. The shrinkage parameter $\lambda$ is~$0.1$.

@@

 The \textsc{mnist} dataset with the observation size $K=100$ is utilised. We use the training data of size $50$, the pool data of size $500$,  and the test data of size $100$. The algorithm learns on the training data and it is evaluated on the test data. To actively collect a next data point from the pool, the algorithm is used to predict a point with the highest uncertainty. The selected point is moved from the pool to the training data and the algorithms learns on the updated training data. Overall, $10$ pool additions are performed. After every addition the performance is measured on the test data. We compare the active approach of adding new points from the pool with the random approach that picks a new data point from the pool at random. The procedure is repeated for $20$ times.
